Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-04-18 manuscript/manuscript.tex
Improved linear density text in manuscript

On the homogeneous density tesseroids case, \citet{Uieda2016} and 
\citet{Grombein2013} compared their numerical model with the analytical 
solution of a spherical shell.
We could follow this idea, but for our needs the spherical shell must 
have the same density function as the numerical model, although the 
analytical solution in this case must be derived.

@@

The integrals that define the gravity fields in equations 
\ref{eq:tesseroid-pot}-\ref{eq:tesseroid-tensor} have analytical 
solutions in case of a spherical shell with homogeneous density 
\citep{Mikuska2006,Grombein2013}.
But for variable density spherical shell the problem wasn't address in 
the literature yet.

@@

Lets consider a linear density function like:

@@

\noindent The gravitational potential generated by a spherical shell 
with this density on any external point at distance $r$ from its centre 
can be obtained from the following expression:

@@

When applying the density-based discretization algorithm on any 
tesseroid with linear density, the absolute difference defined in 
equation \ref{eq:density-abs-diff} will always be zero, so the 
discretization inequation \ref{eq:delta-density} will never be 
satisfied.
Therefore, independently of the linear function proposed, the 
density-based algorithm will never be applied, leaving the modified 
adaptive discretization algorithm as the only mechanism to control the 
accuracy of the numerical approximation.
That's why we only need to determine the minimum value of $D$ needed in 
order to guarantee an acceptable accuracy while ignoring the value of 
$\delta$.

@@

We have computed these gravitational fields on every point of the four 
grids for different values of $D$, ranging from 0.5 to 10 with a step 
of 0.5.
Then we calculated the maximum difference between these results and the 
value that the analytical solution assumes at the same height.
Because the gravity fields generated by a tesseroid are approximated by 
the effect of point masses, the numerical results may vary between 
computation points at same height but at different latitude or 
longitude.
That's why we choose for comparison the maximum difference between the 
results on every point of the grid and the analytical solution.

@@

In order to perform the comparison, we proposed a thin and a thick 
spherical shells with 1km and 35km of thickness, respectively, whose 
outer radii is equal to the mean Earth radius.
On both cases we will suppose that the density assumes a value of 
2670~kg/m$^3$ on the outer surface and 3300~kg/m$^3$ at the inner 
radius:

@@

Then we computed the percentage difference between the numerical model 
with its corresponding analytical solution (equation 
\ref{eq:density-linear}) for the thick and the thin spherical shell.

We can see the results for the potential, the vertical component of the 
gradient $g_z$ and the $g_{zz}$ component of the Marussi tensor on 
Figures \ref{fig:D-linear-thin} and \ref{fig:D-linear-thick}.
We exclude the results corresponding to the other two diagonal 
components of the tensor because they are similar to the ones obtained 
for the $g_{zz}$, although they can be found on the repository. 
\todo[inline]{chequear}

For the linear density case, we can observe that the calculation of the 
potential and the $g_z$ component of its gradient needs a value of 
$D=1$ and $D=2$, respectively, in order to ensure an accuracy of 0.1\%.
On the other hand, the $g_{zz}$ computations show a noticeable 
difference between the thin and the thick shell: the first one needs a 
value of $D$ equal to 8 while the later only a $D$ of 3.5.
We will keep the more conservative result of $D$ equal to 8 in case of 
computing the Marussi tensor components in case of a linear density 
tesseroid.

>>>>

On the homogeneous density tesseroids case, \citet{Uieda2016} and 
\citet{Grombein2013} compared their numerical models with the analytical 
solution of a spherical shell.
We could follow this idea, but for our needs the spherical shell must 
have the same density function as the numerical model.
In contrast with the homogeneous density case, the analytical solutions 
for a variable density spherical shell must be obtained.

We will then perform these comparisons between the analytical solutions 
and the numerical model for specific density functions: a linear and an 
exponential one.
And from these results we will finally obtain default values for $D$ 
and $\delta$ that ensure a numerical error lower than 0.1\%.

@@

The integrals that define the gravity fields in equations 
\ref{eq:tesseroid-pot}-\ref{eq:tesseroid-tensor} have almost trivial 
analytical solutions in case of a spherical shell with homogeneous 
density \citep{Mikuska2006,Grombein2013}.
But for a variable density spherical shell the problem haven't been 
addressed in the literature yet.

@@

Lets perform a comparison between our numerical model and the 
analytical solution for a spherical shell with a linear density 
function like the following one:

@@

Firstly we need to solve the integral on equation \ref{eq:shell-pot} in 
order to obtain the analytical solution in this case.
Thus, the gravitational potential generated by a spherical shell with this 
density on any external point at distance $r$ from its centre can be 
obtained from the following expression:

@@

Secondly we can analyse how the numerical algorithm works for any 
tesseroid with linear density.
When applying the density-based discretization algorithm on any 
tesseroid with this kind of density, the absolute difference defined in 
equation \ref{eq:density-abs-diff} will always be zero, so the 
discretization inequation \ref{eq:delta-density} will never be 
satisfied.
Therefore, independently of the linear function proposed, the 
density-based algorithm will never be applied, leaving the modified 
adaptive discretization algorithm as the only mechanism to control the 
accuracy of the numerical approximation.
That's why we only need to determine the minimum value of $D$ needed in 
order to guarantee an acceptable accuracy while ignoring the value of 
$\delta$.

@@

We have computed these gravitational fields on every point of the four 
grids for different values of $D$, ranging from 0.5 to 10 with a step 
of 0.5.
Then we calculated the maximum difference between these results and the 
value that the analytical solution assumes at the same height of the 
computation grid.
Because the gravity fields generated by a tesseroid are approximated by 
the effect of point masses, the numerical results may vary between 
computation points at same height but at different latitude or 
longitude.
That's why we choose for comparison the maximum difference between the 
results on every point of the grid and the analytical solution.

@@

In order to perform the comparison, we proposed a thin and a thick 
spherical shells with 1km and 35km of thickness, respectively, whose 
outer radii are equal to the mean Earth radius.
On both cases we will suppose that the density assumes a value of 
2670~kg/m$^3$ on the outer surface and 3300~kg/m$^3$ at the inner 
radius:

@@

Then we computed the percentage difference between the numerical model 
with its corresponding analytical solution (equation 
\ref{eq:shell-pot-linear}) for the thick and the thin spherical shell.

We can see the results for the potential, the vertical component of the 
gradient $g_z$ and the $g_{zz}$ component of the Marussi tensor on 
Figure \ref{fig:D-linear}.
We exclude the results corresponding to the other two diagonal 
components of the tensor because they are similar to the ones obtained 
for the $g_{zz}$, although they can be found on the repository. 
\todo[inline]{chequear}

For the linear density case, we can observe that the calculation of the 
potential and the $g_z$ component of its gradient needs a value of 
$D=1$ and $D=2$, respectively, in order to ensure an accuracy of 0.1\%.
On the other hand, the $g_{zz}$ computations show a noticeable 
difference between the thin and the thick shell: the first one needs a 
value of $D$ equal to 8 while the later only a $D$ of 3.5.
We will keep the more conservative result of $D$ equal to 8 in case of 
computing the Marussi tensor components.
